Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-03-23 jss/EpiCompare-paper/EpiCompare-paper.tex
A new take on Section 2

\pkg{EpiCompare} delivers \textit{time-invariance} centric analysis by
(1) taking a global, not marginal view of how epidemics move through
time and (2) by treating full epidemics as filaments, not functions
(breaking away from a time-centric view). The following section aims to
highlight the strengths of \textit{time-invariance} and define the
mathematical foundations that \pkg{EpiCompare}'s tools stand upon.

Epidemics are complex objects, and hard to assess and compare to one
another due to the differences in the diseases, the location where the
outbreak occurs, how the effected population reacts, and the time
aspects (including start of the epidemic, speed of infection and more).
Time-invariant analysis attempts to make different epidemics easier to
compare by removing many time dependent aspects of an epidemic and
instead focus on the overall ``life'' of an epidemic, emphasizing the
number of lives affected.

@@

Time-invariance analysis, as it appears in \pkg{EpiCompare}, can be seen
as an attempt to escape many difficulties comparing different epidemics.
With time-invariant analysis, comparing the decades-long outbreak of HIV
in the US to a 10 day outbreak of norovirus on a cruise ship is still
possible. Time-dependent problems can arise when estimation of certain
epidemiological parameters, including the reproductive number \(R_0\),
which \citet{Gallagher2020} showed can be heavily impacted by choices of
beginning and ending time points of an epidemic.

@@

\(R_0\) is probably the most famous time-invariant numerical summary of
an epidemic and is associated with the Susceptible-Infectious-Recovered
(SIR) model. For a demonstration of \(R_0\) getting lost in a
time-dependent analysis we introduce the \citet{Kermack1927}'s common
SIR model. This model captures the transitions from one state to the
next as a system of ordinary differential equations, where \(N\) is the
total number of individuals, \(\beta\) is the rate of infection, and
\(\gamma\) is the rate of recovery,

@@

From this model, \(R_0 = \beta/\gamma\), aka the ratio of the estimated
infection rate compared to the estimated recovery rate.

Interestingly, with regards to traditional epidemic \(state\)
vs.~\(time\) plots, when comparing different epidemics their \(R_0\)s
are often impossible to compare. This can be seen is a basic example,
where we have 2 different epidemics that start in a population of 1000
with 10 individuals infected, but with different parameters related to
equation \ref{eq:ODE}. Specifically suppose that
\(\beta_1, \gamma_1 = (a,b)\) and \(\beta_2, \gamma_2 = (c,d)\) and that
we observe both of these epidemics for 15 days. The epidemic
trajectories are shown in the \(state\) vs.~time plots in Figure
\ref{fig:different-scales-standard}. At a glance, we may assume that
Model 1 has a larger \(R_0\) than Model 2 because the peak of infection
occurs more quickly than in Model 2. On the other hand, we may think
Model 2 has a larger \(R_0\) because we may think the number of
infections in that model has not yet peaked at time 15. Yet both
epidemics have the same \(R_0\) of 2.

However, when we treat both epidemics a functional data through a space
that tells us the proportion of the population each each class, as is
seen in figure 3 we see a different story. The points seem to overlap
and form the same trajectory. Now it seems to be that Model 2 is
following the same trajectory as Model 1 but is not as far along in the
infection process. We can see there is something fundamentally linking
these two different epidemics, and this fundamental link turns out to be
\(R_0\).

We can see this mathematically if we let our two epidemics be presented
as \(\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}\),
\(\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}\) respectively. As with the
example, assume both models have the same initial values
\((S(0), I(0), R(0))\), and let
\(R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}\) where
\(\beta_i\) and \(\gamma_i\) are the average infection rate and recovery
rate, respectively, for SIR model \(i=1, 2\). And define \(a>0\) to be
the relative scalar such that \(\beta_2 = a \beta_1\) if and only if
\(\gamma_2 = a \gamma_1\).

@@

\subsection[Beyond R0 and SIR]{Beyond \(R_0\) and Kerman's and
McKendrick SIR Models}\label{beyond-r0-sir}

Through \(R_0\) we saw that treating epidemics more like a function of
the time points in some higher dimensional space (nicely visualizable
with a ternary plot for the SIR models), then we could better compare
the overall structure of the epidemic and how it impacted the
population. Without knowing the generative structure underlying these
two epidemics, we propose that time-invariant tools can help compare
them in smart ways, and that these ideas can extend past the CITE
kermack1927 model and to epidemics with more states.

@@

The first step of most data analysis is cleaning the data
\textcolor{violet}{so it can \sout{to}} be explored. There are multiple
ways to collect epidemiological data. Sometimes individual records
\textcolor{violet}{are collected}, with times of different states of the
epidemic (infection, recovery, etc.) as well as individual information
like network structure, location, and sub-population
information\textcolor{violet}{.\sout{will be collected, whereas}}
\textcolor{violet}{O}ther data collections focus on aggregate counts of
individuals in each epidemic state. In fact, usually only the number of
new infections at each time step (e.g.~weekly case counts) is observed.
Compartment totals
\textcolor{violet}{(amounts of individuals in each state)} are then
imputed from those case counts along with other information about the
disease and the population of interest.
\textcolor{violet}{In \pkg{EpiCompare}, w}e focus on understanding the
overall impact of an outbreak at the aggregate/population level,
\textcolor{violet}{which allows examination of overall trends of an epidemic more easily}.

@@

\textcolor{violet}{[Ben says: this whole paragraph is new]} In order to
help the practitioner examine epidemics from an aggregate/ population
lens, we provide a function called \texttt{agents\_to\_aggregate}. This
function transforms data about individual/agents' initial entry into
each state (e.g.~start of infection, start of recovery, etc) to an
aggregate view of how many individuals where in a state at a given time.
There are often situations where grouping agents into subpopulations
(e.g.~subpopulations defined by age or sex) can highlight different
aggregate level trends. For example, research by
\citet{rvachev1985,anderson1992,worby2015} develop state-based models
that account for subpopulations in their
models\footnote{\textcolor{violet}{[Ben says: I don't really like this sentence, but I'm trying to include the references - are they needed?]}}.
This function allows the user to easily group agents into new
subpopulations by using the \pkg{tidyverse} paradigm to first
\texttt{group\_by()}\footnote{The \texttt{group\_by} function is from
  the \pkg{dplyr} package.} the subpopulation defining features and then
aggregate (using \texttt{agents\_to\_aggregate}) conditional on these
subpopulations.
\textcolor{orange}{I think these last two sentences are repetitive}

\textcolor{violet}{[Ben says: this whole paragraph is new]} The
\texttt{agents\_to\_aggregate} function is also flexible to deal with a
wide range of information about each individual. It can, theoretically,
account for infinitely many states, allowing the practitioner to
aggregate information relative to the standard ``Suspectible'',
``Infectious'', and ``Recovered'' states (CITE) or also add states like
``Exposed'', ``iMmune'', ``Hospitalized'' and more. Additionally,
\texttt{agents\_to\_aggregate} also permits indicators for death/exit
and birth/entry dates (that change the overall population size).
\textcolor{orange}{Don't think this is necessary \sout{This function is the highest tested function in this package, and can deal with many nuances that might come-about in the individual data.}}
The function \texttt{agents\_to\_aggregate()} is a powerful tool in the
step of pre-processing data, one that quickly and simply transforms
individual level information into an aggregate view that can be
processed and analyzed in a simpler manner.

@@

\textcolor{violet}{[Ben says: this is actually an older version:]} With
raw data, ``getting to know'' our data currently means figuring out good
combinations of visualizations, numerical summaries and subsets. An
expert coder can start from \texttt{agents\_to\_aggregate} to
successfully do this in many ways, but \pkg{EpiCompare} also includes
tools to rapidly explore your data if your has at least three epidemic
states (although these tools work best with exactly three epidemic
states). Our \texttt{geom\_aggregate} provides a rapid way to explore
different subpopulations' experience of the epidemic, and combines the
ideas behind \texttt{agents\_to\_aggregate} for the SIR case, with
\texttt{geom\_path} and \texttt{coord\_tern} to visualize any number of
groups epidemic trajectory in 3d simplex space using \pkg{ggplot2} and
\pkg{ggtern} \citep{Wickham2016, Hamilton2018}. Visualization tools for
SIR models were developed because (1) SIR models are some of the most
common and basic epidemic state-based models and (2) our simplex
representation of these epidemics emphasizes a ``time-invarance''
representation of the data (for a refresher see Section
\ref{sec:time-invariant}).

@@

\textbf{Model Fitting and Simulations}
\textcolor{violet}{[Ben says: see footnote 3 about the introduction to this paragraph.]}

Although this package does not focus on fitting a model to data, we do
provide some flexible functions for simulation of basic discrete-time
epidemic-state
models.\sout{ with Bernoulli or Multinomial transitions which can be used to (noisily) approximate ODEs which describe transitions from one state to the next}\footnote{The
  original sentence seemed clearer to me (it may even be a run-on
  sentence now). I'm wondering if you didn't like the original sentence
  relative to the ``truth-ful-ness'' of the statement? Here's the
  original sentence slightly edited: ``Although this package does not
  focus on estimating a model for the data, we do provide some power
  functions for simulation of basic state models with Bernoulli
  approximation of ODE state models encoded in simulate\_agents and
  simulate\_SIR\_agents''.}. These functions can be naturally combined
with \texttt{agents\_to\_aggregate()} to proceed in the pipeline. The
function \texttt{simulate\_SIR\_agents()} simulates an SIR epidemic with
user inputs for the number of simulations, the initial number in each
state, the infection and recovery parameters \((\beta, \gamma\)), and
the total number of discrete time steps. This function allows for easy
access to SIR model analysis and comparison. Beyond SIR models, the
function \texttt{simulate\_agents()} takes as input a user-specified
transition matrix and other epidemic parameters to allow the user to
create simulations of an outbreak for \textit{any} number of states and
any number of transitions among them. This allows for users to explore
the space of models in an intuitive way without getting bogged down by
too much mathematical detail. For consistency, we have made output from
\texttt{simulate\_agents()} and \texttt{simulate\_SIR\_agents()}
compatible with \texttt{agents\_to\_aggregate()} so aggregate
information may easily be accessed.

>>>>

\textcolor{orange}{time invariance vs. time invariant analysis??}

\textcolor{orange}{I again changed this section a lot, but hopefully it's more in line with what you are thinking.  The story I want to tell is 1) introduce the concept of time invariance; 2) demonstrate time invariance through R0 and tell why it is important; 3) visualize R0 from the SIR model with a time-invariant plot; and 4) talk about models beyond SIR.}

\textcolor{orange}{I noticed the ternary plot kinda fell out of this section and it's not really detailed anywhere else in the paper.  So I put it back in.  It's a non-standard plot and a feature of EpiCompare so I think it's worth talking about, and I think it's appropriate to introduce in this section.}

\pkg{EpiCompare} delivers \textit{time-invariance} centric analysis by
(1) taking a global, not marginal view of how epidemics move through
\sout{time} \textcolor{orange}{populations} and (2) by treating full
epidemics as
filaments\sout{ not functions (breaking away from a time-centric view)}
and not points produced by functions of time. The following section aims
to highlight the strengths of \textit{time-invariance} and define the
mathematical foundations that \pkg{EpiCompare}'s tools stand upon.

\textcolor{orange}{Mathematically}, epidemics are complex objects. They
are hard to assess and compare to one another due to the differences in
the diseases, the location where the outbreak occurs, how the
\textcolor{orange}{a?}ffected population reacts, and the timefeatures
(including start of the epidemic, speed of infection and more).
Time-invariant analysis \sout{attempts to make}
\textcolor{orange}{makes} different epidemics easier to compare by
removing many time dependent aspects of an epidemic. Instead
\textcolor{orange}{it} focuses on the overall \sout{"life"}
\textcolor{orange}{trajectory} of an epidemic, emphasizing the number of
lives affected.
\textcolor{orange}{really like the lives affected part; I think that's a powerful statement}

@@

Time-invariance analysis, as it appears in \pkg{EpiCompare},
\sout{can be seen as an attempt to escape} \textcolor{orange}{bypasses}
many difficulties \textcolor{orange}{in} comparing different epidemics.
With time-invariant analysis, comparing the decades-long outbreak of HIV
in the US to a 10 day outbreak of norovirus on a cruise ship is
\sout{still} possible. Time-dependent problems can arise when estimation
of certain epidemiological parameters, including the reproductive number
\(R_0\), which \citet{Gallagher2020} show can be \sout{heavily impacted}
\textcolor{orange}{strongly effected} by choices of beginning and ending
time points of an epidemic.

@@

\(R_0\) is probably the most famous time-invariant numerical summary of
an epidemic and is \textcolor{orange}{often} associated with the
Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2001}. For a
demonstration of \(R_0\) getting lost in a time-dependent analysis we
introduce the \citet{Kermack1927}'s common SIR model. This model
captures the transitions from one state to the next as a system of
ordinary differential equations, where \(N\) is the total number of
individuals, \(\beta\) is the rate of infection, and \(\gamma\) is the
rate of recovery,

@@

From this model, \(R_0 = \beta/\gamma\), aka the ratio of the estimated
infection rate compared to the estimated recovery rate. Since \(\beta\)
and \(\gamma\) are both rates (of infection and recovery, respectively),
the ratio of the two, \(R_0\), is a time-invariant quantity.
\textcolor{orange}{Once $R_0$ is estimated, then important epidemic quantities can be inferred such as a the percent of a population needed to be vacccinated to stop the sustained spread of an epidemic and the total number of infections, the final size.  Moreover, $R_0$ allows us to compare different disease outbreaks and different outbreaks on the same scale.  }

\subsection[visualizing r0]{Visualizing \(R_0\)}\label{vis-r0}

With regards to traditional epidemic \(state\) vs.~\(time\) plots, when
\sout{comparing}\textcolor{orange}{examining} different epidemics their
\(R_0\)s are often impossible to compare. This can be seen in a basic
example, where we have two different epidemics that start in two
populations of 1000 people with 10 individuals
\textcolor{orange}{initially} infected, but with
\sout{different parameters}
\textcolor{orange}{different infection and recovery rates (see Eq. \ref{eq:sir-ODE}}).
Specifically suppose that \(\beta_1, \gamma_1 = (a,b)\) and
\(\beta_2, \gamma_2 = (ca,cb)\) where \(a, b, c > 0\). Further suppose
we observe both of these epidemics for 15 days. The epidemic
trajectories are shown in the \(state\) vs.~time plots in Figure
\ref{fig:different-scales-standard}. At a glance, we may \sout{assume}
\textcolor{orange}{believe} that Model 1 has a larger \(R_0\) than Model
2 because the peak of infection occurs more quickly than in Model 2. On
the other hand, we may \sout{think} \textcolor{orange}{believe} Model 2
has a larger \(R_0\) because we may think the number of infections in
that model has not yet peaked at time 15. Yet, both epidemics have the
same \(R_0\) of 2.

\textcolor{orange}{Introduce ternary plots here.  The below paragraphs are mostly new}

However, time-invariant centric analysis allows us to visualize \(R_0\)
for the Kermack and McKendrick SIR model. We call the epidemic
trajectory of number of individuals in each state a \textit{filament}.
Mathematically, a filament \(\psi\) for the SIR model \(\psi\) this is
the set of points of the number in each state for each time point, \[
\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t=0:T}.
\]

Since the number in each state is non-negative and the sum over the
three states for a given time point sums to \(N\), then \(\psi\) will
lay in a 2d triangular plane, i.e.~a ternary plot. As a result, we can
visualize the global trajectory of an epidemic in a single 2d plot and
ultimately \(R_0\). \textcolor{orange}{Show pic here?}

We visualize the two epidemics in a global, ternary view in Figure
\ref{fig:different-scales-tern}. Without getting into too much detail of
the intricacies of this plot, we immediately see the points of the two
filaments \(\psi\) seem to form the same trajectory. Now, it is much
clearer that Model 2 is following the same trajectory as Model 1 but is
not as far along in the infection process. We can see there is something
fundamentally linking these two different epidemics, and this
fundamental link turns out to be \(R_0\).

We can show this mathematically if we let our two epidemics be presented
as \(\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}\),
\(\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}\) respectively. As with the
example, assume both models have the same initial values
\((S(0), I(0), R(0))\), and let
\(R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}\) where
\(\beta_i\) and \(\gamma_i\) are the average infection rate and recovery
rate, respectively, for SIR model \(i=1, 2\). And define \(a>0\) to be
the relative scalar such that \(\beta_2 = a \beta_1\) if and only if
\(\gamma_2 = a \gamma_1\).

@@

{\centering \includegraphics{Figs/unnamed-chunk-3-1}

\caption{\label{fig:different-scales-tern}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.  Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.  These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered.}\label{fig:unnamed-chunk-3}
\end{figure}
\end{CodeChunk}

\subsection[Beyond R0 and SIR]{Beyond \(R_0\) and Kermack's and
McKendrick SIR Models}\label{beyond-r0-sir}

Through \(R_0\) \textcolor{orange}{example}, we see that treating
epidemics more like a
\sout{function of the time points in some higher dimensional space (nicely visualizable with a ternary plot for the SIR models)}
\textcolor{orange}{filaments embedded in a lower dimensional space},
then we can better compare the overall structure of the epidemic and
\sout{how it impacted the population}
\textcolor{orange}{see how the population was directly impacted}.
\sout{Without knowing the generative structure underlying these two epidemics},
\textcolor{orange}{When the underlying generative model is unknown for epidemic trajectories},
we propose that time-invariant tools can help compare them in smart
ways, and that these ideas can extend past the CITE kermack1927 model
and to epidemics with more states.

@@

The first step of most data analysis is cleaning the data
\textcolor{violet}{so it can \sout{to}} be explored. There are multiple
ways to collect epidemiological data. Sometimes individual records
\textcolor{violet}{are collected}, with times of different states of the
epidemic (infection, recovery, etc.) as well as individual information
like network structure, location, and sub-population
information\textcolor{violet}{.\sout{will be collected, whereas}}
\textcolor{violet}{O}ther data collections focus on aggregate counts of
individuals in each epidemic state. In fact, usually only the number of
new infections at each time step (e.g.~weekly case counts) is observed.
Compartment totals
\textcolor{violet}{(amounts of individuals in each state)} are then
imputed from those case counts along with other information about the
disease and the population of interest.
\textcolor{violet}{In \pkg{EpiCompare}, w}e focus on understanding the
overall impact of an outbreak at the aggregate/population level,
\textcolor{violet}{which allows for examination of overall trends of an epidemic more easily}.

@@

\textcolor{violet}{[Ben says: this whole paragraph is new]} In order to
help the practitioner examine epidemics from an aggregate/ population
lens, we provide a function called \texttt{agents\_to\_aggregate}. This
function transforms data about individual/agents' initial entry into
each state (e.g.~start of infection, start of recovery, etc.) to an
aggregate view of how many individuals were in a state at a given time.
There are often situations where grouping agents into subpopulations
(e.g.~subpopulations defined by age or sex) can highlight different
aggregate level trends. For example, research by
\citet{rvachev1985,anderson1992,worby2015} develop state-based models
that account for subpopulations. In \pkg{EpiCompare}, we facilitate
subpopulation analysis by combining \pkg{dplyr}::\texttt{group\_by} and
\texttt{agent\_to\_aggregate} to provide aggregation at a group level.

\textcolor{violet}{[Ben says: this whole paragraph is new]} The
\texttt{agents\_to\_aggregate} function is flexible and can deal with a
wide range of information about each individual. It can, theoretically,
account for infinitely many states. This functionality allows the
practitioner to aggregate information relative to the standard states
(e.g.~``Susceptible'', ``Infectious'', and ``Recovered'') as well as add
states (e.g.~``Exposed'', ``iMmune'', ``Hospitalized'') (CITE).
Additionally, \texttt{agents\_to\_aggregate} also permits indicators for
death/exit and birth/entry dates. Overall, the function
\texttt{agents\_to\_aggregate()} is a powerful tool for pre-processing
data.

@@

\textcolor{violet}{[Ben says: this is actually an older version:]} With
raw data, ``getting to know'' our data currently means figuring out good
combinations of visualizations, numerical summaries and subsets. An
expert coder has many ways to successfully explore the data in an
aggregate lens using \texttt{agents\_to\_aggregate}. \pkg{EpiCompare}
also includes tools to rapidly explore data that has three epidemic
states. Building on the tools in \pkg{ggplot2} and \pkg{ggtern}, our
\texttt{geom\_aggregate} provides a rapid way to explore different
subpopulations' experience of an epidemic by combining the ideas behind
\texttt{agents\_to\_aggregate} for three-state models to examine
subpopulation trajectories in a 3d simplex space
\citep{Wickham2016, Hamilton2018}. \textcolor{orange}{come back?}
Visualization tools for three-state models were developed because (1)
SIR models are some of the most common and basic epidemic state-based
models and (2) our simplex representation of these epidemics emphasizes
a ``time-invarance'' representation of the data (for a refresher see
Section \ref{sec:time-invariant}).
\textcolor{orange}{make sure SIR is defined before}

@@

\textbf{Model Fitting and Simulations}

Although this package does not focus on fitting a model to data, we do
provide some flexible functions for simulation of basic discrete-time
epidemic-state models. These functions produce individual level
information and can be naturally combined with
\texttt{agents\_to\_aggregate()} to view these simulations through an
aggregate lens. The function \texttt{simulate\_SIR\_agents()} simulates
an SIR epidemic with user inputs for the number of simulations, the
initial number in each state, the infection and recovery parameters
\((\beta, \gamma\)), and the total number of discrete time steps. This
function allows for easy access to SIR model analysis and comparison.
Beyond SIR models, the function \texttt{simulate\_agents()} takes as
input a user-specified transition matrix and other epidemic parameters
to allow the user to create simulations of an outbreak for \textit{any}
number of states and any number of transitions among them. This
flexibility in states can be used to also reflect group-based dynamics.
This allows for users to explore the space of models in an intuitive way
without getting bogged down by too much mathematical detail. For
consistency, we have made output from \texttt{simulate\_agents()} and
\texttt{simulate\_SIR\_agents()} compatible with
\texttt{agents\_to\_aggregate()} so aggregate information may easily be
accessed.
